---
title: "Study 12 — cross-aspect synthesis"
subtitle: "SBayesRC truth versus SBayesR-LV truth"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r}
root_candidate <- if (file.exists(file.path(getwd(), "DESCRIPTION")))
  getwd() else file.path(getwd(), "..", "..")
root <- normalizePath(root_candidate, winslash = "/", mustWork = TRUE)
tables <- file.path(root, "results", "local",
  "12_logvar_annotation_evaluation", "aspect2", "tables")
performance <- utils::read.csv(file.path(tables,
  "cross_aspect_performance.csv"))
convergence <- utils::read.csv(file.path(tables,
  "cross_aspect_convergence.csv"))
decision <- utils::read.csv(file.path(tables, "aspect2_decision.csv"))
```

## Reciprocal design

Aspect 1 fits R/LV/RC to the canonical SBayesRC-generated truth. Aspect 2 fits
the same three implementations to SBayesR-LV-generated truth while holding
the genotype, people, marker order, annotations, LD blocks, retained-rank rule,
h2, chain schedule, and validation framework fixed.

```{r}
knitr::kable(performance, digits = 4)
knitr::kable(convergence, digits = 3)
```

## Interpretation

```{r}
knitr::kable(decision)
```

The synthesis asks whether model ranking changes with the generating
annotation mechanism and where differences attenuate from annotation
architecture through PIP/components, beta, and prediction. Two single
simulation scenarios cannot support a universal model ranking.

Model ranking is architecture-dependent in only one direction in these two
single realizations. Under RC truth, RC clearly has the best beta recovery and
validation genetic-value correlation, with LV between RC and R. Under LV
truth, however, LV and R are effectively tied, and RC remains within the
registered downstream equivalence margin. Aspect 2 therefore does not produce
the symmetric own-model advantage that Outcome B anticipated.

The attenuation hierarchy is consistent across aspects. Annotation and
component architectures differ strongly; PIP rankings differ more than beta;
posterior beta and validation prediction are much closer. LV theta sampling is
substantially better behaved than RC alpha sampling in both aspects, but in
Aspect 2 the stable LV q posterior still falls short of the preregistered q
truth-recovery screen. These two scenarios support mechanism-sensitive
inference, not a general model-superiority claim.
