---
title: "Study 06A: Annotation-model estimability"
description: "Closed at EST-R2: stable probability rankings and SNP inference with uncertain annotation-contrast magnitudes."
---

```{r}
#| echo: false
source("R/benchmark-reporting.R")
capsule <- "results/reference/06_annotation_models/current-stop"
rcsv <- function(name) read.csv(file.path(capsule, name), check.names = FALSE)
design <- rcsv("annotation_design_summary.csv")
fit_status <- rcsv("fit_status.csv")
computation <- rcsv("computational_summary.csv")
```

::: {.callout-important}
## Current status

**CLOSED — EST-R2.** Probability/ranking functions are stable, but
annotation-contrast magnitudes remain uncertain.

```text
Study 06 status: CLOSED
Primary final decision: EST-R2
Official qualifier: EST-R5
Standard unrestricted continuous-alpha sampler development: closed at PMA-R3
Additional final benchmark / scientific fit: not required
Next methodology: separate Bayesian annotation-selection / annotation-PIP study
```

The pinned official trajectory provides descriptive evidence, but formal
independently seeded multichain replication remains unavailable under the
v0.2.6 native RNG contract. Historical failed qualifications and intermediate
decisions below remain part of the audit trail; they are not the current study
status.
:::

## Final authoritative scientific result

Functional annotations produce reproducible SNP-prior rankings and stable
SNP-level inference, while the unrestricted continuous-alpha decomposition and
exact annotation-effect magnitudes are substantially less reliable. Enriched
and continuous effects are directionally stable; the null annotation is
compatible with zero; magnitude is quantitatively uncertain and route-sensitive.
This is **EST-R2**. The authoritative synthesis is the
[final conclusion](../../../docs/studies/study06_final_conclusion.md), with detailed
evidence in the
[estimability analysis](../../../docs/studies/study06_estimability_and_contrasts.md)
and identity in
[`final_decision.json`](../../../results/reference/06_annotation_models/final_decision.json).

## What Study 06 tests

Study 06 asks whether correctly specified annotations improve causal-marker
prioritization and genetic-signal recovery in BayesRC/SBayesRC relative to
matched BayesR/SBayesR, and whether the learned annotation hierarchy, marker
allocation, and BED versus summary-statistic likelihood routes are reliable.

The main scientific comparison is individual-level BED BayesRC versus retained
block-eigen SBayesRC. The latter uses the prespecified block factorization and
all 100 positive modes in every block at `eigen_prop = 0.995`; this tests the
retained-factor route, block ordering, summary-model semantics, and blockwise
likelihood factorization, not substantial eigenspace truncation.

The immutable v2 identities are:

```text
Specification:      241c15afab8fefc571e38e625130de6e4ab58b958c30cff369e26998ee30fa56
Informative truth:  169d52bef390022a9106d7e61b200493869b40cbb76f1c8d911eebbc80fea1eb
```

## Designs and formal decisions

### V1 sparse design — failed and preserved

V1 used about 37,991 chromosome-1 markers, 1,400 training people, and about 50
expected non-null markers with active weights 0.60/0.30/0.10. That implies only
about 30/15/5 expected markers in the three active components. BED histories
failed convergence and approximate sparse CSR failed its residual-scale
contract. The tracked `current-stop` capsule is immutable failed-development
evidence, not a completed benchmark or a method-performance comparison.

```{r}
#| echo: false
knitr::kable(fit_status, caption = "Historical v1 fit status (historical evidence).")
knitr::kable(computation, digits = 3,
  caption = "Historical v1 recorded computation (descriptive).")
```

V1 remains useful conceptually as `v1_sparse_stress`: a late-stick, prediction,
PIP, active-count, variance, and numerical-integrity stress scenario. It is not
the qualification of precise late-stick annotation recovery.

### V2 identifiable design — formal qualification failed

V2 uses 2,000 people (1,400 training, 600 validation), 1,500 markers in 15
separated 100-marker blocks, heritability 0.50, and
`gamma = c(0, 0.01, 0.1, 1)`. The informative truth has component counts
1,329/84/50/37, hence 171 non-null markers. Informative and marginally matched
uninformative annotation scenarios use intercept, enriched binary, continuous
signal, and null columns. Deterministic truth, annotation-rank, stick-cell,
marker-order, and block audits all passed.

All four registered learned BayesRC/SBayesRC qualification entries completed,
but every entry failed the multichain convergence contract. Informative
later-stick directions and BED/block heritability agreement also failed. At
that historical gate this was a formal qualification failure and the then-
envisioned final benchmark was not authorized. The later estimability analysis,
not an additional fit, supplied the final EST-R2 endpoint.

### Large information-scale feasibility — completed, learned joint mixing failed

The separate one-replicate profile froze all 5,000 canonical individuals and
37,991 QC markers. Its annotation, truth, stick-identifiability, GWAS, and 76
full-positive-mode block audits passed; realized counts were
36,791/618/392/190 and realized heritability was exactly 0.50. The historical
phase stopped as `LARGE-F6`, which remains documented. After compact aggregate
traces and the GCTB-compatible block residual contract were validated, all six
frozen fits completed. E0/B0 and fixed-alpha E2/B2 pass aggregate latent and
variance diagnostics; learned E1/B1 fail joint alpha, occupancy, and active-
count convergence. SNP ranking remains useful, and the BED/block h2 offset
persists. The primary decision is `LARGE-G2`, with `LARGE-G3/G4` secondary. See
the [final hierarchy of evidence](../../../results/reference/06_annotation_models/final_hierarchy_of_evidence.csv)
and [final decision](../../../results/reference/06_annotation_models/final_decision.json).

| Fit | Core convergence | AUPRC | realized active | Vg | route h2 | g-truth cor. |
|---|---|---:|---:|---:|---:|---:|
| E0 BED baseline | pass | .151 | 1,072 | .713 | .349 | .835 |
| B0 block baseline | pass | .134 | 1,143 | 1.054 | .515 | .838 |
| E2 fixed-alpha BED | pass | .304 | 1,088 | .761 | .372 | .866 |
| B2 fixed-alpha block | pass | .305 | 1,142 | 1.049 | .513 | .859 |
| E1 learned BED | fail; descriptive | .312 | 144 | .719 | .352 | .834 |
| B1 learned block | fail; descriptive | .254 | 1,093 | 1.048 | .512 | .855 |

Block `ves` is retained mean block residual variance, whereas BED `ves` is
global individual-level residual variance. Route h2 uses each registered route
definition; direct residual-variance equality is neither expected nor tested.

## Evidence levels

The rest of this report uses three levels:

- **Established** — supported by converged comparisons, exact tests, or a
  clearly delimited validated descriptive comparison whose scope is stated.
- **Descriptive but useful** — pooled nonconverged summaries, one official
  deterministic trajectory, within-path drift, uncontrolled runtime, or final
  values without histories.
- **Unresolved** — competing explanations cannot yet be separated or required
  diagnostics are unavailable.

“Expected active count” means the sum of marker-specific prior non-null
probabilities. “Realized active count” or “component occupancy” means sampled
marker allocations. They are not interchangeable.

## Current conclusions

### Established

- BayesR/SBayesR baselines converge.
- Fixed-true-alpha BayesRC/SBayesRC converges for the retained quantities, with
  strong cross-chain allocation agreement where complete occupancy histories
  were unavailable.
- Audited stick eligibility, Albert–Chib latent variables, alpha conditionals,
  intercept handling, `sigmaSqAlpha` conditional, and component-probability
  transformation are mathematically and computationally correct.
- The alpha hierarchy converges rapidly when allocations are frozen.
- Fixing `sigmaSqAlpha` and replacing the teaching prior by the
  production-equivalent annotation-variance prior do not restore dynamic joint
  convergence.
- Repeating hierarchy updates improves BED alpha marginal diagnostics, but the
  full occupancy/variance posterior still fails. Repeating allocation sweeps is
  unfavorable.
- The tested complete-state coupling ladder passes exact tiny-state checks but
  has effectively zero Study 06 overlap. Retained lower-replica state is
  insufficient to evaluate exact partial exchange.
- The roughly 0.09–0.10 BED/block heritability offset persists without
  annotation learning, with true alpha fixed, and with every block mode
  retained. It is distinct from learned-alpha feedback and eigen truncation.
- Informative annotations improve causal-marker ranking. The official SBayesRC
  trajectory shows the same prioritization benefit descriptively on this truth.
- Matched official SBayesRC and `sblr` strongly agree for SNP effects, PIPs,
  ranking, and validation genetic values on the delimited single trajectory.
- In the registered 5,000-person/37,991-marker replicate, baseline and
  fixed-true-alpha aggregate quantities converge, but both ordinary learned-
  alpha routes fail joint alpha/allocation convergence. Increasing information
  scale alone did not resolve the bottleneck in this replicate.

### Descriptive but useful

The learned-alpha AUPRC and prediction summaries below are descriptive under
non-convergence. Official alpha/architecture summaries, native five-component
results, visible drift, and runtime are single-trajectory descriptive. Pooled
learned BED/block summaries and per-chain implementation comparisons are not
formal posterior validation.

| Fit | Evidence | AUPRC | AUROC | expected active | h2 | effect–truth cor. | validation-g cor. | prediction cor. |
|---|---|---:|---:|---:|---:|---:|---:|---:|
| BayesR BED baseline | formally converged | .3975 | .6943 | 167.8 | .4036 | .8922 | .8932 | .6314 |
| SBayesR block baseline | formally converged | .3270 | .6679 | 178.9 | .5050 | .8602 | .8698 | .6033 |
| learned informative BED | descriptive under non-convergence | .5948 | .8528 | 58.7 | .4162 | .9138 | .9178 | .6545 |
| learned informative block | descriptive under non-convergence | .5407 | .8300 | 131.7 | .5093 | .8847 | .8956 | .6316 |
| fixed-alpha BED | fixed-alpha, converged available contract | .6075 | .8628 | 171.0 | .4273 | .9242 | .9267 | .6616 |
| fixed-alpha block | fixed-alpha, converged available contract | .5993 | .8551 | 179.5 | .5134 | .8909 | .8994 | .6350 |
| BED H20 | nonconverged joint state | .5945 | .8520 | about 55.6 | .4164 | .9138 | .9178 | .6546 |
| block H20 | nonconverged joint state | .5419 | .8309 | about 133.6 | .5090 | .8850 | .8958 | .6320 |

### Difficult or delimited at closure

- Official SBayesRC multichain convergence and quantitative alpha parity.
- Quantitative parity of official `sigma_anno` and `sblr` `sigmaSqAlpha`.
- Latent architecture, component occupancy, and active/component-count
  semantics across implementations.
- Residual-variance and effect/`nbsq` scale contracts.
- BED versus block-eigen variance and heritability calibration.
- Unrestricted continuous-alpha joint learning, global alpha/allocation
  coupling, and later-stick quantitative interpretation. The same-posterior
  development line is closed at PMA-R3 rather than left as a current roadmap.

## Post-closure isolated alpha recovery

The **ALPHA-R1 — isolated standard SBayesRC alpha hierarchy validated** addendum
does not reopen Study 06 or change EST-R2. With fixed latent `z`, scalar Gibbs
reproduced the exact Gaussian posterior (maximum mean error 0.00428; maximum
covariance error 0.00084). With known continuation outcomes, blocked and scalar
chains agreed closely (maximum mean difference 0.00936; maximum R-hat 1.00127).

Across 20 independent simulated hierarchies and four blocked-update chains per
fit, maximum mean R-hat was 1.00074, maximum individual R-hat was 1.00199, and
95% coverage ranged from .90 to 1.00. Enriched-slope posterior SD increased from
.118 to .198 to .279 over the three sticks, but its repeated biases were only
-.017, -.002, and -.012. This is increasing finite-sample uncertainty, not
systematic conditional alpha failure.

Eligible counts averaged 1,500, 170.7, and 84.2; enriched eligible counts
averaged 149.1, 77.2, and 42.3. Informative enriched SNPs therefore became about
45% and 50% of later eligible sets rather than remaining 10%. Total information
declines, while selective continuation partly preserves information about the
informative slope.

```text
fixed z: exact alpha conditional
known outcomes: calibrated, well-mixed alpha inference
full SBayesRC: difficult joint alpha <-> allocation/occupancy exploration
```

PMA-R3 is consistent with this diagnosis: it supplied an exact coordinated
global move under the same posterior, but was computationally impractical. The
2024 Jian Zeng R-reference total-`m` versus eligible-`n` intercept detail changed
the isolated stick-2/stick-3 enriched means only modestly (.505/.508 and
-.475/-.442); it is a source-contract note, not an explanation of the full
problem or a claim about production GCTB. Compact evidence and the standalone,
non-production reproduction script are registered in
[`alpha_recovery_metadata.json`](../../../results/reference/06_annotation_models/alpha_recovery_metadata.json).
- Exact annotation-contrast magnitudes and some hierarchy-scale quantities,
  despite stable directions and downstream SNP outputs.

## Implementation and sampler lessons

Package-side experiments are clearly separated from benchmark-side experiments.
Low-level detail remains authoritative in the linked `sblr` documents.

| Implementation question | Finding | Scientific implication | Status | Authoritative evidence |
|---|---|---|---|---|
| Shared annotation implementation | Scalar BED, CSR, and block-eigen routes call the shared annotation-prior kernel; route adapters were audited directly. | Similar route behavior is not merely assumed from output. | correctness established | [`sblr` hierarchy audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_alpha_hierarchy_joint_sampling_audit.md) |
| Stick eligibility/orientation | Stick 1 uses all markers; later sticks use continuers. Truth eligible/continue/stop counts are 1500/171/1329, 171/87/84, 87/37/50. Native and independent references agree. | The observed late-stick sample sizes and continuation convention are known exactly. | correctness established | same |
| Albert–Chib and alpha conditionals | Latent truncation, residualized Gaussian means/variances, and proper intercept handling agree to numerical precision. | No audited conditional-formula defect explains mixing. | correctness established | same |
| `sigmaSqAlpha` conditional | The intercept is excluded; three non-intercepts enter the inverse-chi-square conditional. Current and production-equivalent priors were mapped and exactly checked. | Conditional correctness does not establish joint convergence. | correctness established | same |
| Component probabilities | Sequential stick transformation, flooring, normalization, and orientation passed numerical checks; ordinary states were not floor-controlled. | Numerical flooring is not the Study 06 cause. | correctness established | same |
| BED component traces | Trace markers were resolved in the full-Glist index domain instead of the fitted subset. Boundary resolution/validation was corrected without changing transitions or RNG order. | Retained traces are now reliable; the defect did not cause learned-alpha mixing. | defect corrected | same |
| Frozen allocations | Current prior, production prior, fixed variance, and fixed-alpha reference all pass when allocations are fixed to truth. | Isolated hierarchy conditionals are not the primary failure. | candidate explanation ruled out | same; decision C |
| Fixed `sigmaSqAlpha` | Dynamic fits still fail on both routes. | A centred annotation-variance funnel is not the primary bottleneck. | candidate explanation ruled out | same |
| Production-prior ablation | The production-equivalent prior does not restore dynamic convergence. | Prior-tail choice alone is insufficient. | candidate explanation ruled out | same |
| S1/H5/H20/A5/A20 | H20 improves BED alpha marginals, but joint occupancy/variance still fails; allocation repetition worsens behavior. | More conditional work alone is not the missing global transition. | candidate explanation ruled out | [`sblr` K4 audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_allocation_hierarchy_kernel_composition.md) |
| Coupling tempering | Tiny detailed balance/endpoints pass, but Study 06 accepts 0/2,400 exchanges with replicas hundreds of log-density units apart. | Algebraically correct tempering is an ineffective bridge for this posterior. | candidate explanation ruled out | [`sblr` T4 screen](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_bed_coupling_tempering_screen.md) |
| Partial exchange | Exact formulas exist, but lower-replica complete states were not retained; no proposal was validated or rejected. Aggregate active-count mismatch is not the sole explanation. | The candidate could not be evaluated from existing state. | unresolved contract difference | [`sblr` F6 audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_partial_exchange_feasibility.md) |
| BED/block variance calibration | The h2 offset persists without annotation, with fixed alpha, and with 100/100 modes. | Route calibration is distinct from alpha feedback and eigen truncation. | unresolved contract difference | paired isolation; package audits |
| Block runtime/storage | Operators are constructed before MCMC, BED is not reread during MCMC, and immutable block operators are shared. Recorded block runs were faster; analytical and observed process memory differ. | The retained block route is computationally attractive, but recorded times are not controlled benchmarks. | performance observation | [`sblr` block contract](https://github.com/psoerensen/sblr/blob/master/docs/dev/blr_block_eigen_contract.md) |
| Official/`sblr` contracts | Residual variance, effect/`nbsq` scale, p1 orientation, pi updates, active-count definitions, component counts, `sigma_anno`/`sigmaSqAlpha`, and wrapper labels differ or remain unmapped. | Strong SNP agreement does not validate latent architecture. | unresolved contract difference | [final decision](../../../results/reference/06_annotation_models/final_decision.json) |

## Official SBayesRC comparison

The external implementation is **official SBayesRC R implementation v0.2.6,
source SHA `b95d3fcbad8ff358290922a58fff893439296138`**. “Official SBayesRC” below
means that implementation; it is not used as a generic synonym for GCTB.

### Strongest matched comparisons

| Comparison | PIP Pearson/Spearman | effect Pearson | validation-g correlation | top-50/top-100 | AUPRC |
|---|---:|---:|---:|---:|---:|
| Official D1 vs learned block `sblr` | .957/.964 | .998 | .998 | 46/86 | .557 vs .541 |
| Official D1 vs learned BED `sblr` | .933/.953 | .956 | .960 | 39/77 | .557 vs .595 |

Official D1 resembled every saved learned `sblr` chain, rather than one isolated
occupancy regime: block PIP correlations were .946–.965 and BED correlations
.925–.936; effect correlations were .99831–.99833 and .95483–.95664.

Official annotation benefit was large on the identical truth: D0 versus D1
AUPRC was .320 versus .557, AUROC .664 versus .842, validation genetic-value
correlation .869 versus .896, and phenotype prediction .604 versus .634.

> The `sblr` SBayesRC implementation compares well with official SBayesRC for
> SNP effects, PIPs, causal prioritization, validation prediction, and practical
> runtime. Agreement is weaker or unresolved for latent architecture,
> later-stick alpha coefficients, component occupancy, and annotation-variance
> parameters.

Official SBayesRC v0.2.6 does not connect its public seed argument to the native
RNG streams, so multichain convergence parity remains unavailable. The official
results are single-trajectory descriptive and must not be called converged.

## Alpha and annotation variance

Only unambiguously mapped values are compared. The v2 non-intercept truth for
stick 1 is enriched 1.60, continuous .30, null 0. The calibrated truth
intercept is retained in the simulation object but is not duplicated in the
compact result documents; no value is inferred here.

| Quantity, stick 1 | Truth | Official D1 single trajectory | `sblr` learned BED pooled | `sblr` learned block pooled | Fixed-alpha |
|---|---:|---:|---:|---:|---:|
| intercept | available in truth object | -2.406 | not compactly reported here | not compactly reported here | fixed to registered truth |
| enriched | 1.600 | 1.507 | 1.897 | 1.642 | 1.600 |
| continuous | .300 | .586 | .638 | .622 | .300 |
| null | 0 | -.156 | -.042 | -.371 | 0 |
| evidence label | truth | single-trajectory descriptive | pooled nonconverged | pooled nonconverged | fixed-alpha |

Official first-stick enriched and continuous directions agree with truth and
every saved learned `sblr` chain; its null coefficient is comparatively small.
Quantitative coefficient equality is not established. BED H20 put all 12 alpha
marginals within the diagnostic contract, but it remained a nonconverged joint
state; the compact package evidence reports diagnostics rather than a complete
mapped mean table.

Later-stick values are not quantitatively interpretable. Official p3 intercept
drifted about 1.85 posterior SD and p3 continuous about -1.06 SD; `sblr`
later-stick alpha and occupancy also mixed poorly.

Official final `sigma_anno` values were 1.976/1.043/4.803 for D1 and
3.480/.384/.413/.605 for D2. `sblr` informative selected-window
`sigmaSqAlpha` means were about 2.14/1.90/3.14 BED and 2.04/1.54/1.79 block;
these are pooled nonconverged posterior summaries. In production-prior dynamic
fits, BED stick-specific `sigmaSqAlpha` diagnostics passed while block sticks
1/2 failed; this did not rescue the joint posterior. Fixed-variance conditions
set `sigmaSqAlpha = 1` and are ablations, not posterior estimates.

Official retained `sigmaSqAlpha` histories are not exposed. Official
`sigma_anno` and `sblr` `sigmaSqAlpha` must not be assumed identical without a
source-contract audit, and final official values cannot establish posterior
mean, uncertainty, drift, ESS, or chain parity.

## Derived annotation functions and final estimability result

### Why stick 1 is scientifically central

With four components, stick 1 is the active-versus-null decision:

\[
P(c_i>0\mid A_i)=q_{i1}=\Phi(A_i\alpha_1).
\]

Later sticks redistribute already-active SNPs among nonzero components. The
frozen truth has 1,500/171/87 eligible SNPs across sticks 1/2/3. The enriched
column occurs in exactly 225/1,500 SNPs (15%), and in 107/171 and 62/87 of the
later eligible sets. The information loss is therefore the sharp reduction in
eligible latent continuation outcomes; no claim that only a handful of enriched
eligible SNPs remain is needed. This helps explain why first-stick informative
directions can be useful while later-stick decomposition remains weak.

### Counterfactual active/non-null contrasts

| Annotation | Truth | `sblr` BED | `sblr` block | Official D1 descriptive |
|---|---:|---:|---:|---:|
| enriched binary | .423 | .192 [.081, .397] | .299 [.119, .642] | .204 [.117, .302] |
| continuous signal | .0886 | .0555 [.0239, .104] | .112 [.0383, .236] | .077 [.042, .118] |
| null annotation | 0 | -.007 [-.068, .046] | -.068 [-.268, .060] | -.024 [-.096, .024] |

The shared ordering is enriched much greater than continuous, with null near
zero. Directions and rankings are convincing; exact magnitudes are biased,
route-sensitive, and incompletely covered. The continuous example makes the
distinction concrete: raw first-stick values are truth .300, official .586,
BED .638, and block .622, while the induced active contrasts are .0886, .077,
.0555, and .112. The probability-level functional is more biologically useful
and closer to truth, but not perfectly calibrated. The raw null values similarly
map to smaller contrasts whose intervals include zero.

## Why SBayesRC can work despite difficult alpha parameters

The isolated-alpha addendum establishes that alpha estimation works when the
hierarchy is known. The full learned model is harder because alpha changes SNP
mixture probabilities and the resulting allocations/effects determine the next
alpha regression. Local conditionals can be correct while this coupled joint
state is traversed slowly. Later sticks are less precise, but are not shown to
be intrinsically unidentified.

For stick \(k\),

\[
\eta_{ik}=A_i\alpha_k,\qquad q_{ik}=\Phi(\eta_{ik}),
\]

and the four mixture probabilities are

\[
\pi_{i0}=1-q_{i1},\quad
\pi_{i1}=q_{i1}(1-q_{i2}),\quad
\pi_{i2}=q_{i1}q_{i2}(1-q_{i3}),\quad
\pi_{i3}=q_{i1}q_{i2}q_{i3}.
\]

```text
alpha
→ A alpha
→ q
→ SNP-specific pi
→ latent components/effects
→ SNP PIP / posterior beta
→ genetic values / prediction
```

Raw alpha is an upstream latent parameterization. Correlated coefficient
configurations—such as lower intercept plus larger enriched slope—can behave
similarly over the realized annotation profiles. The full-rank annotation
matrix (rank 4, condition number about 2.23) rules out a simple design null
space, so the supported language is weak practical determination and strong
posterior dependence, not formal non-identifiability.

The probit map can compress latent differences in saturated regions:
`Phi(4)` is about .99997 and `Phi(7)` is effectively 1. Official D1 provides a
striking descriptive example: half-specific `A alpha` vector correlation is as
low as -.975 while the minimum nonconstant induced probability-vector Pearson
correlation is at least .999. Its raw-alpha ESS range is about 1.4–86.2 and the
largest half shift about 1.25 posterior SD. These are halves of one native
trajectory, not independent chains.

Posterior inference also averages over the hierarchy:

\[
P(c_i>0\mid y)=E_{\alpha,\theta}
\left[P(c_i>0\mid y,\alpha,\theta)\right].
\]

Thus no single plug-in alpha value is the ordinary SNP endpoint. There is no
contradiction between difficult alpha inference and useful SBayesRC SNP
inference: the induced probability functions can be substantially more stable,
and posterior averaging can yield reproducible marginal PIPs and effects. Poor
MCMC exploration is not harmless in general; this is the empirical hierarchy
observed on the frozen fixture.

### Three levels of estimability and reporting

| Level | Study 06 behavior | Reporting role |
|---|---|---|
| Latent alpha / `sigmaSqAlpha` | weak or correlated in some directions, especially later sticks | secondary plus R-hat, ESS, MCSE, intervals, covariance |
| `A alpha`, q/pi, contrasts, rankings | substantially more stable; direction strong, magnitude uncertain | primary annotation-derived functions; `A alpha` secondary |
| SNP PIP, beta, genetic value, prediction | highly stable | primary SNP and downstream outputs |

Probability ranking is reproducible: active-prior Spearman is at least
.986/.950 and largest-component-prior Spearman at least .999/.995 for BED/block.
PIP Pearson is at least .994/.969, signed-beta Pearson .9999/.9991, and
genetic-value chain correlation .99994/.99925. BED chain ranges are AUPRC
.591–.597, AUROC .849–.856, genetic-value truth correlation .917–.918, and
prediction .654–.655; block ranges are .514–.560, .818–.840, .896–.897, and
.631–.633.

R-hat near one is agreement, not truth recovery. Expected active count averages
about 351.8 BED and 807.4 block versus truth 180 despite R-hat near one and very
large variability. Therefore convergence is not calibration, and stable ranking
is not exact probability-magnitude recovery.

Future SBayesRC reports should emphasize SNP PIPs/effects and downstream
performance; annotation-induced active/component probabilities, rankings, and
counterfactual contrasts with uncertainty; and retain raw alpha, `A alpha`,
hierarchy scales, covariance, R-hat, ESS, and MCSE as secondary diagnostics.
Raw alpha can inform first-stick direction but should not be the sole biological
summary.

## Runtime observations

| Method | Trajectories/chains | Iterations | Wall time |
|---|---:|---:|---:|
| Official D0 SBayesR | 1 | 9,000 | 4.61 s |
| Official D1 matched SBayesRC | 1 | 9,000 | 39.28 s |
| Official D2 native SBayesRC | 1 | 9,000 plus tuning | 82.03 s |
| `sblr` informative block SBayesRC | 4 | 9,000 each | 54.74 s |
| `sblr` informative BED BayesRC | 4 | 9,000 each | 129.24 s |

Official runs used one OpenMP and one BLAS thread. `sblr` used four registered
chains and may use parallel execution; retained tracing, diagnostics, package
builds, and process overhead differ. This is descriptive, not a controlled
performance benchmark.

> The `sblr` block-eigen route is computationally competitive. Four `sblr`
> block chains required only about 1.4 times the wall time of one official
> matched trajectory in the recorded runs.

## Experiment timeline

| Phase | Question and route/repository | Result | Implication | Evidence |
|---|---|---|---|---|
| 1. v1 sparse qualification | Sparse BED/CSR annotation qualification | Failed; preserved | Too sparse for primary late-stick qualification; CSR fidelity confounded | formal qualification |
| 2. v2 identifiable design | Increase active/stick information | Deterministic gates passed | Accidental empty truth cells ruled out | design/protocol |
| 3. v2 qualification | Learned BED/block BayesRC/SBayesRC | All fits completed; all failed convergence | Final benchmark historically blocked at this gate | formal qualification |
| 4. paired power isolation | Baseline, learned, shuffled, fixed-alpha | Baselines/fixed pass; learned fails; ranking benefit large | Alpha/allocation feedback isolated | diagnostic experiment |
| 5. mathematical audit | Shared conditionals in `sblr` | Exact checks pass | Formula defect ruled out | correctness established |
| 6. frozen allocations | Hierarchy with truth allocations | Converges rapidly | Isolated hierarchy ruled out | exact/converged diagnostic |
| 7. variance ablations | Fixed variance and production prior | Dynamic failure persists | Variance funnel/prior not primary | diagnostic experiment |
| 8. trace correction | BED selected-marker index domain | Defect corrected; RNG/transitions unchanged | Reporting defect not mixing cause | defect corrected |
| 9. kernel schedules | S1/H5/H20/A5/A20 | H20 helps alpha; joint fails; A repeats worsen | Recomposition alone insufficient | K4 diagnostic |
| 10. tempering | Tiny exact plus Study 06 BED | Algebra passes; 0 exchanges | Tested ladder ineffective | T4 diagnostic |
| 11. partial exchange | Offline retained-state audit | Exact assessment unavailable | Candidate neither validated nor rejected | F6 unresolved |
| 12. official export/API | Exact Study 06 official input and seed test | Export valid; independent chains blocked | GCTB-P5 remains | formal decision |
| 13. official single path | D0/D1/D2, 9,000 iterations | Strong SNP agreement/benefit; architecture drift | Descriptive parity completed | single-trajectory descriptive |
| 14. documentation consolidation | This task | Evidence hierarchy normalized | One authoritative reading path | documentation decision |
| 15. block contract validation | small official D0/D1 vs new S0/S1 | SNP/residual gates passed | `gctb_block` semantics validated without relabeling history | SMALL-G1 |
| 16. larger feasibility | `n=5000`, `m=37,991` | controls pass; learned joint mixing fails; SNP utility strong | information scale does not remove alpha/allocation bottleneck; route offset persists | LARGE-G2/G3/G4 |
| 17. exact transition audits | pair, coordinated subset, PX, particle Gibbs, particle-marginal alpha | exact references; no scalable qualification solution | standard same-posterior development endpoint reached | AA-R5 / PMA-R3 |
| 18. estimability and contrasts | frozen 1,500-marker retained chains and official D1 | rankings/SNP outputs stable; contrast magnitude uncertain | final scientific closure | EST-R2; EST-R5 qualifier |

## Closure and next model

Study 06 is complete. No further unrestricted continuous-alpha sampler
engineering or additional Study 06 scientific fit is recommended. The
same-posterior development line ended at PMA-R3: exact but computationally
impractical. On this controlled benchmark, `sblr` reproduces the main behavior
of official SBayesRC—similar first-stick directions, broadly comparable active
contrasts, highly concordant SNP inference, and a qualitative separation
between difficult alpha and stable induced probabilities—without claiming
universal posterior or implementation equivalence.

The next methodology is Bayesian annotation selection / annotation PIPs as a
distinct statistical model and study. It targets annotation relevance rather
than acting as a sampler fix for standard SBayesRC.

## Navigation and reproducibility

Start with the [final conclusion](../../../docs/studies/study06_final_conclusion.md),
then the [estimability evidence](../../../docs/studies/study06_estimability_and_contrasts.md)
and this report. Historical experiment-specific ledgers remain available in
Git history; their current scientific content is retained here and in those
two authoritative documents. The final closure identity is
[`final_decision.json`](../../../results/reference/06_annotation_models/final_decision.json).
This report reads only the unchanged v1 capsule and never launches a sampler.

The 06A analysis entry point validates compact accepted evidence only. It
invokes no sampler, regenerates no truth, and requires no retained fit object.
